---
title: "STAT 812 - Assignment 1"
author: "Kangjie Zhang"
date: "October 16, 2018"
output:
  md_document:
    variant: markdown_github
    df_print: paged
    toc: yes 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 1.1
### function: data_gen_fm

```{r}
#use method of inverse cdf to generate iid sample from exp(1/2)
gen_exp0.5<-function(n)
{
  u<-runif(n)
  -2*log(1-u)
}

#use method of inverse cdf to generate iid sample from normal(0,1)
gen_normal<-function(n)
{
  n1<-ceiling(n/2)
  theta<-runif(n1,0,2*pi)
  R<-sqrt(gen_exp0.5(n1))
  x<-R*sin(theta)
  y<-R*cos(theta)
  c(x,y)[1:n]
}

#generate a data set from factor analysis model: (Y,X)=Z*T+E
#T: Given parameters, k*(p+1) matrix
#Z: Data of the factors, n*k matrix
#E: Noise term, n*(p+1) matrix
data_gen_fm<-function(T,n,p,k)
{
  if (n<=0 || p<=0 || k<0)
  {stop("Invalid argument, not satisfy n>0, p>0 or k>=0")
  }
  if (k>0)
  {
    Z<-matrix(gen_normal(n*k),n,k)
    E<-matrix(gen_normal(n*(p+1)),n,(p+1))
    YX<-Z%*%T+E
  }else{
    E<-matrix(gen_normal(n*(p+1)),n,(p+1))
    YX<-E
  }
  
  list(Y=YX[,1],X=YX[,-1],YX=YX)
}
```


### function: pv_lm_select
```{r}
#Select p_star features and then fits a linear model, return p-value based on F-statistic
#YX: Data set
#p_star: the number of selected features
pv_lm_select<-function(YX,p_star)
{
  if (p_star<=0)
  {stop("Invalid argument")
  }
  
  cor<-abs(cor(YX[,1],YX))
  ix<-sort(cor,decreasing = TRUE, index.return = TRUE)$ix[1:(p_star+1)]
  d<-data.frame(YX[,ix])
  Y<-d[,1]
  d1<- data.frame(Y,d[,-1])
  fit.lm<-lm(d1$Y ~ .,data = d1)
  fvalue<-summary(fit.lm)$fstatistic
  
  1-pf(fvalue[1],fvalue[2],fvalue[3])
}

#simulations: get a p-value with T=0
N<-2000;n<-40;p<-1000;p_star<-20;k<-0
pv_lm_select(data_gen_fm(0,n,p,0)$YX,p_star)
```

### Explain:
The p-value for the model after selecting p_star features is very small, so the model is significant. On the other hand, the selected features have the hightest correlation with y,  which explains why the model is significant.

### function: test_pv_lm_select
```{r}
#use data_gen_fm to generate N data sets;
#use pv_lm_select to compute the p-values; 
#draw histogram of these N p-values
test_pv_lm_select<-function(N,n,p,p_star)
{
  A<-array(0,dim = c(n,p+1,N))
  pv<-rep(0,N)
  for (i in 1:N) {
    A[,,i]<-data_gen_fm(0,n,p,0)$YX
    pv[i]<-pv_lm_select(A[,,i],p_star)
  }
    hist(pv,xlab="2000 pvalues for n=40, p=1000, p_star=20")
}

#simulations
N<-2000;n<-40;p<-1000;p_star<-20;k<-0
test_pv_lm_select(N,n,p,p_star)
```

### Explain:
The p-values for the model after selecting p_star features with N data sets are very small, so the models are significant (because the selected features have the largest absolute correlations with y). It violates the uniformity of p-values of fitting a linear model with selected variables.


## Question 1.2
```{r}
#no_perm: number of permutations for y
#permute y no_perm times; 
#for each permuted data set, get the p-value from lm
pv_lm_select_perm<-function(YX,p_star,no_perm)
{
  if (p_star<=0 || no_perm<=0)
  {stop("Invalid argument")
  }
  pv1<-pv_lm_select(YX,p_star)
    
  index <- 0
  for (i in 1:no_perm) {
    YX2<-data.frame(sample(YX[,1]),YX[,-1])
    pv2 <- pv_lm_select(YX2,p_star)
    if ( pv2 <= pv1){
      index <- index+1 
    }
  }
  (index+1)/(no_perm+1)
}

#simulations
N<-2000;n<-40;p<-1000;p_star<-20;k<-0;no_perm<-100
pv_lm_select_perm(data_gen_fm(0,n,p,0)$YX,p_star,no_perm)
  

#use data_gen_fm to generate N data sets;
#use pv_lm_select_perm to compute the p-values; 
#draw histogram of these N p-values
test_pv_lm_select_p<-function(N,n,p,p_star,no_perm)
{
  A<-array(0,dim = c(n,p+1,N))
  pv<-rep(0,N)
  for (i in 1:N) {
    A[,,i]<-data_gen_fm(0,n,p,0)$YX
    pv[i]<-pv_lm_select_perm(A[,,i],p_star,no_perm)
  }
  hist(pv,xlab="2000 pvalues for n=40, p=1000, p_star=20, no_perm=100")
}

N<-2000;n<-40;p<-1000;p_star<-20;k<-0;no_perm<-100
test_pv_lm_select_p(N,n,p,p_star,no_perm)
```

### Explain:
The p-values from permutation test are uniformly distributed.

## Question 2
### Inverting-CDF method
$$f(x) \propto e^ {l(x)}, \ \ \ f(x) =a e^ {l(x)}$$  

$$
f(x) = \left\{
        \begin{array}{ll}
            ae^{2x+3} & \quad if \ \  x < -1\\
            ae^{2} & \quad if \ \  -1<x \leq 1\\
            ae^{-x+4} & \quad if \ \  x \leq -1\\
        \end{array}
    \right.
$$


$$\int_{ -\infty }^{-1} ae^{2x+3}  \,dx + \int_{ -1}^{1} ae^{2} + \int_{1}^{\infty} ae^{-x+4}  \,dx = 1$$


$$a(\frac{e}{2} + 2e^2 + e^3) = 1 \quad \Rightarrow \quad a=1/(\frac{e}{2} + 2e^2 + e^3)$$
$$\Downarrow$$

#### The CDF of f(x) is:
$$
F(x) = \left\{
        \begin{array}{ll}
            a(\frac{1}{2}e^{2x+3}) & \quad if \ \  x < -1\\
            a(\frac{e}{2}+e^{2}+e^{2}x) & \quad if \ \  -1<x \leq 1\\
            a(\frac{e}{2}+2e^{2}+e^{-x+4}) & \quad if \ \  x \geq 1\\
        \end{array}
    \right.
$$

#### The inverse of F(x) above is:
$$
F^{-1}(u) = \left\{
        \begin{array}{ll}
            [\log(\frac{2u}{a})-3]/2 & \quad if \ \  u<\frac{ae}{2}; \  \quad \quad \quad \quad \quad \quad \quad  \quad (x < -1)\\
            \\
            \frac{u-ae/2}{ae^2}-1 & \quad if \ \  \frac{ae}{2} < u\leq a(\frac{e}{2}+2e^2);  \  \quad \quad  \quad (-1<x \leq 1)\\
            \\
            4-\log(\frac{1-u}{a}) & \quad if \ \ u \geq a(\frac{e}{2}+2e^2); \ \  \quad \quad \quad \ \ \ \quad (x \geq 1)\\
        \end{array}
    \right.
$$

### function: gen_fx
```{r}
#draw samples from f(x) by inverting-CDF method
gen_fx<-function(n)
{
  a<-1/(exp(1)/2+2*exp(2)+exp(3))
  u<-runif(n)
  u1<-subset(u,u<a*exp(1)/2)
  u2<-subset(u,u<a*(exp(1)/2+2*exp(2)) & u>a*exp(1)/2) 
  u3<-subset(u,u>a*(exp(1)/2+2*exp(2)))
  x1<-0.5*log(2*u1/a)-1.5
  x2<-(u2-exp(1)*a/2)/(a*exp(2))-1
  x3<-4-log((1-u3)/a)
  c(x1,x2,x3)
}
```

### function: fx
```{r}
#f(x): piecewise function
fx<-function(x)
{
  a<-1/(exp(1)/2+2*exp(2)+exp(3))
  if (x < -1){
    y <- a*exp(2*x+3)
  }     
  if (x>=1){
    y <- a*exp(-x+4)
  }   
  if (x>=-1 & x<1){
    y <- a*exp(2)
  }
  y
}
```

### Comparison
```{r}
x<-seq(-10,10,by=0.01)
f<-rep(0,length(x))
for (i in 1:length(x)) {
  f[i]<-fx(x[i])
}
#the shape of f(x)
plot(x,f,xlim = c(-10,10),main = "The shape of f(x)")
#histogram of sample draw from gen_fx
hist(gen_fx(100000),xlim = c(-10,10),main = "Histogram of sample(100000) obtained from gen_fx")
```

### Explain:
It can be clearly seen that the histogram of sample (sample size=100000) obtained from my function(gen_fx) has very similar shape as f(x).